---
title: "Reproducible Research - Project 1"
author: "Renga Pandurengan"
date: "2/15/2022"
---

Loading required packages
```{r,  echo=TRUE}

library(knitr)
library(readr)
library(dplyr)
library(ggplot2)

```

#1. Loading & Manipulating data

```{r,  echo=TRUE}

activity <- read_csv("C:/Users/renga/Desktop/Data Science Specialization/Reproducible research/Project1/activity.csv",show_col_types = FALSE)
activity$date_new <- as.Date(activity$date) 

```

```{r,  echo=TRUE}

head(activity)
names(activity)
head(activity)
str(activity)

#total number of missing data
sum(is.na(activity$steps))/dim(activity)[[1]]

#transforming the date column into date format using lubridate
library(lubridate)
activity$date<-ymd(activity$date)
length(unique(activity$date))

```

#2. Creating Histogram of the total number of steps taken each day

```{r, echo=TRUE}
stepsPerDay <- tapply(activity$steps, activity$date, sum, na.rm=TRUE)
qplot(stepsPerDay, xlab='Total steps per day', ylab='Frequency ', main="Total Number of Steps Taken Each Day")

```

#3. Mean and median number of steps taken each day
 
```{r,   echo=TRUE}
 MeanStepsByDay <- mean(stepsPerDay)
MeanStepsByDay
MedianStepsByDay <- median(stepsPerDay)
MedianStepsByDay
```

#4. Time series plot of the average number of steps taken

```{r ,echo=TRUE}
averageStepsPerTimeBlock <- aggregate(x=list(meanSteps=activity$steps), by=list(interval=activity$interval), FUN=mean, na.rm=TRUE)


ggplot(data=averageStepsPerTimeBlock, aes(x=interval, y=meanSteps)) +
  geom_line() +
  xlab("Interval (5 minute)") +
  ylab("average number of steps taken") 
```

#5. The 5-minute interval that, on average, contains the maximum number of steps
```{r,  echo=TRUE}
maxSteps <- which.max(averageStepsPerTimeBlock$meanSteps)
timeMaxSteps <-  gsub("([0-9]{1,2})([0-9]{2})", "\\1:\\2", averageStepsPerTimeBlock[maxSteps,'interval'])
```

#6. Code to describe and show a strategy for imputing missing data


```{r,  echo=TRUE}
#Number of missing values
NoMissingValues <- sum(is.na(activity$steps))
NoMissingValues

#replace missing values  with mean 
activity$steps[is.na(activity$steps)] <- mean(activity$steps, na.rm=TRUE)

```

#7. Histogram of the total number of steps taken each day after missing values are imputed

```{r,  echo=TRUE}
# Create the histogram with imputed data.
qplot(activity$steps, xlab='Total steps per day', ylab='Frequency ')
```

#8. Panel plot comparing  average number of steps taken per 5-minute interval across weekdays and weekends

Are there differences in activity patterns between weekdays and weekends?
 dataset with the filled-in missing values was used.

Creating a new factor variable with two levels - “weekday” and “weekend” (indicating given date is  weekday or weekend ).


```{r,  echo=TRUE}
activity <- mutate(activity, weektype = ifelse(weekdays(activity$date) == "Saturday" | weekdays(activity$date) == "Sunday", "weekend", "weekday"))
activity$weektype <- as.factor(activity$weektype)
head(activity)

# Panel plot 

interval <- activity %>%
  group_by(interval, weektype) %>%
  summarise(steps = mean(steps))
s <- ggplot(interval, aes(x=interval, y=steps, color = weektype)) +
  geom_line() +
  facet_wrap(~weektype, ncol = 1, nrow=2)
print(s)
```



